---
title: "Investigating Variation in Sequencing Depth"
author: Kris Sankaran
date: "`r Sys.Date()`"
---

```{r setup}
library("tidyverse")
library("phyloseq")
library("reshape2")
library("cluster")
library("forcats")

scale_colour_discrete <- function(...)
  scale_colour_brewer(..., palette="Set2")
scale_fill_discrete <- function(...)
  scale_fill_brewer(..., palette="Set2")

theme_set(theme_bw())
min_theme <- theme_update(
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.ticks = element_blank(),
  legend.title = element_text(size = 8),
  legend.text = element_text(size = 6),
  axis.text = element_text(size = 6),
  axis.title = element_text(size = 8),
  strip.background = element_blank(),
  strip.text = element_text(size = 8),
  legend.key = element_blank()
)
```

```{r read_data}
seqtab <- readRDS("../data/seqtab.rds")
taxa <- readRDS("../data/taxa.rds")
survey <- read_csv("../data/WELL_China_1969_7.25.2017.csv")
bc <- readRDS("../data/sample_data_bc.rds")
```

How is sequencing depth related to variables we've measured about the samples?
It looks like it's mainly related to the date during which the survey was taken
(the last three in particular seem to have larger sequencing depths).
```{r depth_variation}
depths <- setNames(rowSums(seqtab), bc$id)
survey$depth <- depths[as.character(survey$id)]
ggplot(survey %>% filter(!is.na(depth))) +
  geom_histogram(
    aes(x = depth)
  ) +
  facet_wrap(~surveydate_pes)
```

This raises two questions

* Are there any species that are more abundant during the different sequencing
  runs (early vs. late samples, in this case)? These might be batch effects.
* Even if there are no batch effects, is a straightforwards clustering of the
  samples just binning people into clusters according to sequencing depth? We
  suspect this might be the case if we don't convert to relative abundances.

To answer the first, we convert species to relative abundances and see if they
are correlated to sequencing depth of the sample from which they came.

```{r depth_abundance_corr}
rel_abund <- seqtab / rowSums(seqtab)
test_stats <- vector(length = ntaxa(seqtab))

high_depth <- depths > 30000
for (j in seq_len(ntaxa(seqtab))) {
  test_stats[j] <- as.numeric(t.test(rel_abund[high_depth, j], rel_abund[!high_depth, j])$statistic)
  if (j %% 50 == 0) {
    message("Processing sequence ", j)
  }
}

hist(test_stats, breaks = 100)
hist(test_stats[abs(test_stats) != 1], breaks = 100)
plot(rel_abund[, which.min(test_stats)], depths)
plot(rel_abund[, which.max(test_stats)], depths)
plot(rel_abund[, order(test_stats, decreasing = TRUE)[2]], depths)
plot(rel_abund[, order(test_stats, decreasing = TRUE)[3]], depths)
qt(0.01, ntaxa(seqtab)) # reference for significance when there's only 1 hypothesis
```
The spikes at 1 and -1 occur when you try to test two samples that are almost
exactly 0. If the first is slightly larger than the second, you get 1, otherwise -1.
I don't really think there's a batch effect, looking at these plots. We've
screened so many RSVs that there are bound to be some where there seems to be an
association, and the number we see seems consistent with it just being chance
variation.

To study the second issue (are our clusters just clustering according to
sequencing depth?), we just do the clustering like we had done before and then
see whether samples with different depths ever appear in the same cluster. The
clustering approach we take here is (1) take asinh transformations (2) reduces
to the top 5 (say) dimensions using MDS with a jaccard distance (you would think
Jaccard distance would be less sensitive to sequencing depth, but remember
higher sequencing depth -> potentially more species detected) and (3) use
partition around medoids in this reduced space. We'll subset to just females, to
be consistent with the approach in earlier analysis.

```{r}
cluster_transformed <- function(x, method = "jaccard", k1 = 5, k2 = 3) {
  D <- phyloseq::distance(asinh(x), method = method)
  cmd_res <- cmdscale(D, k = k1)
  pam(cmd_res, k = k2)
}

f_ix <- bc$gender == "Female"
pam_res <- cluster_transformed(seqtab[f_ix, ])
```

Now we can compare clustering result with sequencing depths.
```{r}
depth_cluster <- data_frame(
  "depth" = depths[f_ix],
  "cluster" = as.factor(pam_res$cluster)
)

ggplot(depth_cluster) +
  geom_boxplot(
    aes(x = cluster, y = depth)
  )
```

We can find those species that have the most variation across clusters. The idea
is to see whether even for these species the clusters are just ordered according
to the overall counts for these species, which might just be a symptom of
differing depths. Since this was done in the MDS-reduced data, we can't just
look at the PAM object's centroids, because those centroids are in the reduced
space. Instead, we first compute centroids in the raw data.

```{r}
x <- asinh(seqtab[f_ix, ])
K <- 3
centroids <- matrix(nrow = ntaxa(seqtab), ncol = K)
for (k in seq_len(K)) {
  centroids[, k] <- colMeans(x[pam_res$cluster == k, ])
}

species_order <- order(
  apply(centroids, 1, var),
  decreasing = TRUE
)

centroids_df <- as_data_frame(
  centroids[species_order, ]
) %>%
  mutate(
    family = fct_lump(taxa[species_order, "Family"], n = 6),
    ix = species_order
  ) %>%
  gather(K, centroid, -family, -ix)

ggplot(centroids_df %>% filter(ix < 200)) +
  geom_tile(
    aes(x = ix, y = K, alpha = centroid, fill = family)
  ) +
  coord_fixed(5) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  theme(
    axis.text.y = element_blank(),
    legend.position = "bottom"
  ) +
  scale_alpha(range = c(0, 1))
```

It looks like, among the 200 species that have the highest variance in cluster
means, the first two rows of clusters have consistently larger read counts than
the bottom row, but between those two, there is some real variation in the
species abundances. This is consistent with the previous plot, which shows that
one of the clusters is just picking up low read count samples. So, while
clustering on the asinh transformed data will find some structure (the
differences between these two high sequencing depth clusters seems real), it
will also pick up on some artifacts, so until a better transformation is found,
it's probably better to use relative abundances.
