---
title: Clustering Counts with Uneven Depths
date: "`r Sys.Date()`"
---

## Approach

There are difficulties in choosing a transformation with which to cluster
microbiome data. Neither of the two usual choices -- relative abundance and
asinh (or log(1 + x)) -- is clearly more applicable than the other,

* asinh: While this preserves the overall sequencing depth structure, it can
  lead to the creation of artificial clusters whose only similarity is that they
  have low sequencing depth.
* relative abundance: While this removes the low-sequencing-depth cluster
  artifacet, it treats all samples the same, when really those samples with
  deeper sequencing have better estimates.

Here, we'll consider alternatives that cluster based on composition, but which
take into account the uneven sequencing depth of different samples. The
essential idea is to model the counts for a single sample as a long multinomial
vector, with $N_{i}$ equal to the total count for that sample and $p_{i}$ equal
to the underlying species composition of that sample. If we allowed each sample
to have a totally different $p_{i}$, then we would be overfitting, and not
sharing any information between / providing any dimension reduction across
samples. Instead, the idea is to impose some structure across the $p_{i}$ --
different strategies yield different methods.

## Setup

But before we do anything complicated, let's load some libraries and read in the
data. The long "theme" text is just to customize the appearance of the ggplot
theme. The centered log transformation is used later for displaying fitted
probabilities (it's like a naive version of PHILR).

```{r setup}
library("tidyverse")
library("rstan")
library("phyloseq")
library("reshape2")
library("forcats")
set.seed(09212017)
knitr::opts_chunk$set(cache = TRUE)

scale_colour_discrete <- function(...)
  scale_colour_brewer(..., palette="Set2")
scale_fill_discrete <- function(...)
  scale_fill_brewer(..., palette="Set2")

theme_set(theme_bw())
theme_update(
  panel.border = element_rect(size = 0.7, fill = "transparent"),
  panel.spacing = unit(0, "cm"),
  panel.grid = element_blank(),
  axis.ticks = element_blank(),
  legend.title = element_text(size = 8),
  legend.text = element_text(size = 6),
  axis.text = element_text(size = 6),
  axis.title = element_text(size = 8),
  strip.background = element_blank(),
  strip.text = element_text(size = 8),
  legend.key = element_blank()
)

#' Centered Log Transform
#'
#' Log transform probabilities for each species and cluster at a time.
centered_log <- function(x) {
  for (i in seq_len(nrow(x))) {
    for (k in seq_len(ncol(x))) {
      x[i, k, ] <- log(x[i, k, ])
      x[i, k, ] <- x[i, k, ] - mean(x[i, k, ])
    }
  }
  x
}
```


Next we read in all the data into a phyloseq object. We're filtering down to
those species that appear in at least 5\% of samples -- this does remove quite a
few species (only 444 are left, out of the original 2565). But to study the
rarer species, I think it would make more sense to analyze the presence /
absence data, and that's not something I try in this note (yet).

```{r read_data}
ps <- phyloseq(
  otu_table = readRDS("../data/seqtab.rds"),
  sample_data = readRDS("../data/sample_data_bc.rds"),
  taxa = readRDS("../data/taxa.rds"),
  phylo = readRDS("../data/phylo_tree.rds")
) %>%
  filter_taxa(function(x) { mean(x > 0) > 0.05}, TRUE) ## appear in 5% of samples

taxa <- tax_table(ps) %>%
  as.data.frame %>%
  rownames_to_column("seq")
phylo_order <- phy_tree(ps)$tip
```

## Dirichlet Multinomial Mixtures

The first approach we'll consider is a Dirichlet Multinomial Model. This model
is like a Gaussian Mixture Model, except instead of modeling cluster component
as a Gaussian, you model it as a Multinomial. The center of each cluster is now
a distinct composition $p_{k}$, and each sample is thought of as a Multinomial
draw from one of these $p_{k}$ (with different depths $N_{i}$).

The code below fits this model with $K = 4$. It's possible to study the analog
of the gap statistic in this model, to choose a better value of $K$, but for now
I'm thinking of $K$ as a tuning parameter rather than any "real-world" quantity
to estimate.

```{r dmm_fit}
K <- 4
stan_data <- list(
  "N" = nsamples(ps),
  "V" = ntaxa(ps),
  "K" = K,
  "x" = get_taxa(ps),
  "alpha" = rep(10, K),
  "gamma" = rep(1e-2, ntaxa(ps))
)

m <- stan_model("dirichlet_multinomial.stan")
dm_fit <- vb(m, stan_data, adapt_engaged = FALSE, eta = 0.1)
dm_res <- extract(dm_fit)
rm(dm_fit) ## free some memory
```

Now we'll study the output. A disadvantage of this model is that it doesn't
directly provide cluster assignments. However, these could be recovered, by
comparing the likelihood of different clusters on the samples that we've
observed. We haven't implemented that here, though.

The main display we consider here shows the fitted probabilities within each of
the $K$ clusters, given in the figure below. Each row is a different cluster,
and each individual boxplot gives the posterior of the centered log probability
for a single species within that cluster (the fact that we're looking at the
entire posterior, rather than just the mean, is why we have boxplots instead of
just points). The species are sorted as in the original OTU table, which is why
the higher probability species appear on the left.

The first thing that stands out is that the first two clusters have basically no
prevotella (the first also is missing a few relatively common ruminoccocus). The
first cluster seems a little more diverse than the others, while the second
seems less diverse -- you can read this by looking at the number of boxplots
above the horizontal 0 reference line.

```{r dmm_inspect}
colMeans(dm_res$theta) ## how common different clusters are

## Join in the taxa data
p_df <- centered_log(dm_res$p) %>%
  melt(
    varnames = c("iter", "k", "v"),
    value.name = "p"
  ) %>%
  mutate(
    seq = taxa_names(ps)[v]
  ) %>%
  left_join(taxa %>% select(seq, Family)) %>%
  mutate(
    family = fct_lump(Family, 7),
    seq = droplevels(factor(seq, levels = phylo_order))
  )

## Plot the fitted probabilities
ggplot(p_df) +
  geom_hline(yintercept = 0, alpha = 0.5) +
  geom_boxplot(
    aes(x = seq, y = p, fill = family, col = family),
    outlier.size = 0.2,
    width = 30
  ) +
  facet_grid(k ~ ., scale = "free_y") +
  labs(y = "clog(p)") +
  theme(
    axis.text.x = element_blank(),
    legend.position = "bottom"
  )
```

## LDA

The main difference between LDA and the Dirichlet Multinomial Mixture model is
that here each sample is allowed to have partial memberships across clusters
(rather than being fixed to any single one). Practically, the big advantage is
that we directly get (partial) cluster assignments (we don't have to do extra
postprocessing, like in the Dirichlet Multinomial).

Formally, instead of assuming that a sample $x_{i}$ is drawn from a multinomial
with underlying probability $p_{k}$ (for one of the clusters $k$), the LDA model
assumes that it's drawn from a mulitnomial with underlying probability $\sum_{k}
\theta_{ik} p_{k}$, where $\theta_{k}$ put different weights on each $p_{k}$.
This let's us estimate something closer to a gradient, rather than total
distinct clusters, since the $\theta_{ik}$'s let us interpolate between
different clusters.

But first, let's fit the model. As before, we set $K = 4$ (without any
particularly good justification).

```{r lda_fit}
m <- stan_model("lda_counts.stan")
lda_fit <- vb(m, stan_data, adapt_engaged = FALSE, eta = 0.1)
lda_res <- extract(lda_fit)
rm(lda_fit)
```

We can now inspect the actual partial cluster assignments ($thetas_{i}$s) and
cluster compositions ($p_{k}$s). The heatmap below shows the posterior mean of
the $\theta_{ik}$'s. The column is a different sample, and each row is a
different cluster. The darkness of a tile is how much that particular cluster
contributed to that sample. We've tried to arrange samples so that those with
similar $\theta_{i}$s appear next to one another. We've also square root
transformed the $theta_{ik}$s, so that differences between smaller probabilities
are easier to see.

The main things to take away here are that most samples have a pretty good
amount of cluster 1, but a few have mostly clusters 3 or 4. A few samples seem
completely dominated by cluster 3. Cluster 2 seems to trade off with cluster 1,
which might mean they represent very similar compositions (one true cluster
getting split into two).

```{r}
theta_hat <- apply(lda_res$theta, c(2, 3), mean) %>%
  melt(
    varnames = c("i", "k"),
    value.name = "theta"
  )

sample_order <- hclust(dist(theta_hat))$order
theta_hat <- theta_hat %>%
  mutate(
    Number = sample_data(ps)$Number[i],
    i = factor(i, levels = sample_order),
    k = as.factor(k)
  )

ggplot(theta_hat) +
  geom_tile(
    aes(x = i, y = k, alpha = sqrt(theta))
  ) +
  coord_fixed(10) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_alpha(range = c(0, 1)) +
  theme(axis.text.x = element_blank())
```

There's a natural follow-up question to this description: Do these clusters have
anything to do with observed phenotype? To study this, we have to compare the
$\theta_{ik}$'s with observed sample information. Below, we give some toy
examples with gender, aoi, and android / gynoid ratio, but these should be
replaced with more meaningful comparisons in the future (either by domain
knowledge or studying correlations separately).

Give one discrete example and two continuous ones. First, we compare the
$\theta_{ik}$s to gender. Each column corresponds to one of the clusters, and
each row is one gender. They seem more or less comparable, though maybe an
argument could be made that the high cluster 3 samples tend to be men, the low
cluster 2 samples tend to be omen, and the low cluster 4 samples tend to be men.
But I'd want to see some separate confirmation (maybe using another method) to
say this confidently.

Instead of comparing against a discrete category (gender), we can compare with
continuous measurements, here aoi and android / gynoid ratios. The value of the
$\theta_{ik}$'s is plotted on the x-axis (on a square root scale), and the
phenotype is on the $y$ axis. The associated topics $k$ are encoded by color.
We've also averaged across each topic using a lowess smooth.

Again, the associations are relatively ambiguous. The only patterns I think are
potentially worth investigating are the decrease phenotypes as membership in
cluster 1 increases and the increase in phenotypes when cluster 3 increases. We
will see when we interpret the topics that this has to do both with diversity
and prevotella, so if these associations held up, then they would be consistent
with the original alpha-diversity story (without using alpha diversity...).

```{r theta_bc_comparison}
theta_supp <- theta_hat %>%
  left_join(sample_data(ps))

## compare with gender
ggplot(theta_supp) +
  geom_histogram(
    aes(x = theta, fill = k)
  ) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_grid(gender ~ k, scale = "free")

## compare with aoi
ggplot(theta_supp) +
  geom_smooth(
    aes(x = theta, y = aoi, col = k),
    alpha = 0.1
  ) +
  geom_point(
    aes(x = theta, y = aoi, col = k),
    alpha = 0.6
  ) +
  scale_x_sqrt() +
  facet_wrap(~gender, scales = "free")

## compare with android gynoid ratio
ggplot(theta_supp) +
  geom_smooth(
    aes(x = theta, y = Android_FM / Gynoid_FM, col = k),
    alpha = 0.1
  ) +
  geom_point(
    aes(x = theta, y = Android_FM / Gynoid_FM, col = k),
    alpha = 0.6
  ) +
  scale_x_sqrt() +
  facet_wrap(~gender, scales = "free")
```

Next we study the composition of the different clusters. The display below is
read in the same way as that for the compositions in the Dirichlet Multinomial
example. Again, we see a prevotella-related difference, with the first cluster
have quite a bit less prevotella and the second having quite a bit more. Cluster
3 seems to have a lot of the rarer species which are not so common in the other
clusters (especially Erysipelos). Cluster 4 seems to have more Ruminoccoceae and
less Bacteroidaceae, while in Cluster 1 that pattern is reversed. Cluster 1
seems somewhat more diverse, though it's a little less clear, since it also has
narrower boxplots (because most samples have some contribution to Cluster 1,
improving the posterior estimates).

```{r beta_hats}
## join in the taxa information
beta_hat <- centered_log(lda_res$beta + 1e-10) %>%
  melt(
    varnames = c("iter", "k", "v"),
    value.name = "beta"
  ) %>%
  mutate(
    seq = taxa_names(ps)[v]
  ) %>%
  left_join(taxa) %>%
  mutate(
    family = fct_lump(Family, 7),
    seq = droplevels(factor(seq, levels = phylo_order))
  )

## plot cluster distribution over species
ggplot(beta_hat) +
  geom_hline(yintercept = 0, alpha = 0.5) +
  geom_boxplot(
    aes(x = seq, y = beta, fill = family, col = family),
    outlier.size = 0.2,
    width = 20
  ) +
  ylim(-9, max(beta_hat$beta) + 0.01) +
  facet_grid(k ~ .) +
  labs(y = "clog(p)") +
  theme(
    axis.text.x = element_blank(),
    legend.position = "bottom"
  )
```

The previous way of plotting the clusters is pretty complicated. We can try to
simplify things by only plotting the mean of each posterior $\theta_{ik}$,
though it's unclear how much this actually helps. The first panel shows the 100
most abundant species, while the second shows the rest (you've got to really
zoom in to see everything though).

The only pattern that seems more notable here than in the boxplot view is the
"competition" between clusters 2 and 3 for Bacteroides and Lachnospiraceae. It
seems that there are quite a few species that are only large in one of these
clusters, but not the other, even though overall both 2 and 3 have quite a few
of these two taxa.

```{r beta_hats_heatmap}
beta_mean <- beta_hat %>%
  group_by(k, v, seq, family) %>%
  summarise(beta_mean = mean(beta))

ggplot(beta_mean %>% filter(v < 100)) +
  geom_tile(
    aes(x = k, y = rev(seq), fill = family, alpha = beta_mean)
  ) +
  coord_fixed(0.1) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_alpha(range = c(0, 1)) +
  theme(
    axis.text.y = element_blank(),
    legend.position = "right"
  )

ggplot(beta_mean %>% filter(v > 100)) +
  geom_tile(
    aes(x = k, y = rev(seq), fill = family, alpha = beta_mean)
  ) +
  coord_fixed(0.1) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_alpha(range = c(0, 1)) +
  theme(
    axis.text.y = element_blank(),
    legend.position = "right"
  )
```
