---
title: Clustering Counts with Uneven Depths
date: "`r Sys.Date()`"
---

```{r setup}
library("tidyverse")
library("rstan")
library("phyloseq")
library("reshape2")
library("forcats")

scale_colour_discrete <- function(...)
  scale_colour_brewer(..., palette="Set2")
scale_fill_discrete <- function(...)
  scale_fill_brewer(..., palette="Set2")

theme_set(theme_bw())
theme_update(
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.ticks = element_blank(),
  legend.title = element_text(size = 8),
  legend.text = element_text(size = 6),
  axis.text = element_text(size = 6),
  axis.title = element_text(size = 8),
  strip.background = element_blank(),
  strip.text = element_text(size = 8),
  legend.key = element_blank()
)
```

```{r read_data}
seqtab <- readRDS("../data/seqtab.rds")
samples <- readRDS("../data/sample_data_bc.rds")
taxa <- readRDS("../data/taxa.rds")
```

Let's test out the dirichelt multinomial mixture model code (on just a subset of
samples).

```{r }
filter_thresh <- 400
stan_data <- list(
  "N" = nsamples(seqtab),
  "V" = filter_thresh,
  "K" = 4,
  "x" = as.matrix(seqtab)[, seq_len(filter_thresh)],
  "alpha" = rep(1, 4),
  "beta" = rep(1, filter_thresh)
)

m <- stan_model("dirichlet_multinomial.stan")
vb_fit <- vb(m, stan_data)
cluster_res <- extract(vb_fit)
```

We can study the output -- this method doesn't actually give cluster
assignments, but it gives cluster compositions, from which we can usually guess
the cluster assignments.

```{r}
colMeans(cluster_res$theta) ## how common different clusters are
p_hat <- apply(cluster_res$p, c(2, 3), mean)
p_df <- p_hat %>%
  melt(
    varnames = c("k", "v"),
    value.name = "p"
  )
p_df$seq <- taxa_names(seqtab)[p_df$v]

## include sequence map in tax table
taxa <- taxa %>%
  as.data.frame %>%
  rownames_to_column("seq")

p_df <- p_df %>%
  left_join(taxa) %>%
  mutate(family = fct_lump(Family, 7))
ggplot(p_df %>% filter(v < 100)) +
  geom_tile(
    aes(x = v, y = k, alpha = sqrt(p), fill = family)
  ) +
  scale_alpha(range = c(0, 1))
```

So it looks like the fourth cluster has more prevotella. The others can be
summarized based on a few species, but it doesn't really seem consistent across
the whole family.
